Analysis paper

### 2021-06-01 analysis/analysis.Rmd
create and streamline plots and tables

```{r}
dailytruth_data %>%
  dplyr::filter(target_end_date >= min(forecast_dates$full_hub_period), 
                target_end_date <= max(forecast_dates$full_hub_period)) %>%
  ggplot(aes(y = true_value, x = target_end_date)) + 
  geom_point() + 
  geom_line() + 
  facet_grid(target_type ~ location_name, scales = "free")

@@

## Visualisation of 2 week ahead forecasts and scores

Todo:

-   do we want to keep the decreasing / increasing points at all on the plot?
-   it is a bit odd that here we look at the classification at the target end date, while below we look at it in terms of the forecast_date

```{r truth-vs-forecasts, fig.height=10, fig.width=15}
# load unfiltered combined data 
df <- unfiltered_data[grepl("2", target) & 
                      quantile %in% c(0.025, 0.25, 0.75, 0.975)]
df <- dcast(df, ... ~ quantile, value.var = "prediction")

# add trend classification by target_end_date (and remove classification by forecast_date)
df <- merge(
  df[, c("classification", "speed") := NULL], 
  copy(epitrend)[, forecast_date := NULL], 
  by = c("target_end_date", "location_name", "target_type")
)

truth_and_obs_plot <- function(df) {
  p <- df %>%
  ggplot(aes(y = true_value, x = target_end_date)) + 
  geom_linerange(aes(ymin = `0.25`, ymax = `0.75`, color = model), 
                 size = 1.5,
                 alpha = 1,
                 position = position_dodge(width = 6)) + 
  labs(y = "value", x = "date") + 
  geom_point(aes(shape = classification)) + 
  geom_line() + 
  # geom_rect(aes(xmin = min(forecast_dates$christmas) + 14, # 2wk ahead forecast
  #               xmax = max(forecast_dates$christmas) + 14,
  #               ymin = -Inf,
  #               ymax = Inf), fill = "lightgrey", alpha = 0.002) +
  theme_light() + 
  facet_wrap(~ location_target, scales = "free", ncol = 2) + 
  theme(legend.position = "bottom") + 
  theme(panel.background = element_rect(fill = "#e1f0fe")) + 
  scale_x_date(limits = c(min(forecast_dates$full_hub_period) + 14, 
                          max(forecast_dates$full_hub_period)))
  return(p)
}

truth_vs_forecast_2_cases <- df %>%
  dplyr::filter(target_type == "case") %>%
  truth_and_obs_plot() + 
  scale_color_manual(values = brewer.pal(name = "Accent", n = 5)[2:5])

ggsave(here("analysis", "plots", "truth-and-2-wk-ahead_cases.png"), 
       plot = truth_vs_forecast_2_cases)

truth_vs_forecast_2_deaths <- df %>%
  dplyr::filter(target_type == "death") %>%
  truth_and_obs_plot()  + 
  scale_color_manual(values = brewer.pal(name = "Accent", n = 5)[1:5]) + 
  annotate('rect', xmin = min(forecast_dates$full_hub_period) + 14,
                xmax = max(forecast_dates$death_not_scored) + 14,
                ymin = 0,
                ymax = Inf, fill = "grey10", alpha = 0.1)

ggsave(here("analysis", "plots", "truth-and-2-wk-ahead_deaths.png"), 
       plot = truth_vs_forecast_2_deaths)

@@

```{r scores-over-time, fig.height=10, fig.width=15}
# load unfiltered combined data 
df <- unfiltered_data[grepl("2", target)]

# plot with scores over time ---------------------------------------------------
scores_target_forecastdate <- eval_forecasts(
  data = df,
  summarise_by = c("model", "location_name", "target_type", 
                   "forecast_date", "target_end_date", "location_target", 
                   "classification"), 
  compute_relative_skill = TRUE, 
  baseline = "Baseline"
)

scores_over_time_plot <- function(scores) {
  p <- scores %>%
  ggplot(aes(x = as.Date(target_end_date), y = scaled_rel_skill)) + 
  geom_line(aes(colour = model)) + 
  geom_point(aes(shape = classification, colour = model)) + 
  theme_light() + 
  facet_wrap( ~ location_target, ncol = 2, scales = "free_y") +
  theme(legend.position = "bottom") +
  labs(y = "relative skill", x = "date") + 
  # geom_rect(aes(xmin = min(forecast_dates$christmas) + 14,
  #               xmax = max(forecast_dates$christmas) + 14,
  #               ymin = 0,
  #               ymax = Inf), fill = "lightgrey", alpha = 0.002) +
  theme(panel.background = element_rect(fill = "#e1f0fe")) +
  scale_y_log10() + 
  scale_x_date(limits = c(min(forecast_dates$full_hub_period) + 14, 
                          max(forecast_dates$full_hub_period)))
  return(p)
}

rel_skill_time_2_case <- scores_target_forecastdate %>%
  dplyr::filter(target_type == "case") %>%
  scores_over_time_plot() + 
  scale_color_manual(values = brewer.pal(name = "Accent", n = 5)[2:5])

ggsave(here("analysis", "plots", "rel-skill-over-time-cases.png"), 
       plot = rel_skill_time_2_case)

rel_skill_time_2_deaths <- scores_target_forecastdate %>%
  dplyr::filter(target_type == "death") %>%
  scores_over_time_plot() + 
  scale_color_brewer(palette = "Accent") + 
  annotate('rect', xmin = min(forecast_dates$full_hub_period) + 14,
                xmax = max(forecast_dates$death_not_scored) + 14,
                ymin = 0,
                ymax = Inf, fill = "grey10", alpha = 0.1)

ggsave(here("analysis", "plots", "rel-skill-over-time-deaths.png"), 
       plot = rel_skill_time_2_deaths)
```

```{r fig.height=10, fig.width=15}
# glue two plots together

# cases_plot <- truth_vs_forecast_2_cases / rel_skill_time_2_case + 
#   plot_layout(guides = "collect") & 
#   theme(legend.position = 'bottom') 
# 
# deaths_plot <- truth_vs_forecast_2_deaths / rel_skill_time_2_deaths + 
#   plot_layout(guides = "collect") & 
#   theme(legend.position = 'bottom') 
truth_vs_forecast_2_cases / rel_skill_time_2_case / truth_vs_forecast_2_deaths / rel_skill_time_2_deaths + 
  plot_layout(guides = "collect") & 
  plot_annotation(tag_levels = 'A') &
  theme(legend.position = 'bottom')

# for some reason the A / B annotation doesn't work here

ggsave(here("analysis", "plots", "figure-forecasts-2.png"))
```

@@

```{r}
# create helper function to make table
make_score_table <- function(scores) {
  scores <- as.data.table(scores)
  setnames(scores, 
           old = c("model", "target_type", "interval_score", "aem", 
                   "relative_skill", "scaled_rel_skill", "sharpness", 
                   "underprediction", "overprediction", "bias", 
                   "50", "90", "location_target", "target_phase"), 
           new = c("Model", "Target", "WIS", "Abs. error", 
                   "rel. skill", "Skill", "Sharpness", 
                   "Underpred.", "Overpred.", "Bias", 
                   "50%-Cov.", "90%-Cov.", "Target", "Target"), 
           skip_absent = TRUE)

@@

  scores <- scores[, .SD,  .SDcols = !c("coverage_deviation", "rel. skill")][
    order(Target, `Skill`)
  ]

  setcolorder(scores, 
              neworder = c("Model", "Target", "Skill"))

  # add an empty line to data.frame to separate deaths and cases
  table <- 
    rbindlist(list(
    scores[1:4], 
    as.list(rep(NA, ncol(scores))), 
    scores[5:9]
  )) %>%
    kable() %>% #format = "markdown"
    kable_styling(bootstrap_options = "striped")
  return(table)
}
```

### 2 weeks ahead

```{r }
# scores by target type for 2 week ahead forecasts

coverage <- eval_forecasts(
  filtered_data[grepl("2", target)], 
  summarise_by = c("model", "target_type", "range"), 
  compute_relative_skill = TRUE, 
  baseline = "Baseline"
)[range %in% c(50, 90), 
                   .(model, target_type, coverage, range)] %>%
  dcast(formula = ... ~ range, value.var = "coverage")

table <- eval_forecasts(
  filtered_data[grepl("2", target)], 
  summarise_by = c("model", "target_type"), 
  compute_relative_skill = TRUE, 
  baseline = "Baseline"
) %>%
  merge(coverage, all.x = TRUE, by = c("model", "target_type")) %>%
  make_score_table()

save_kable(table, file = here("analysis", "plots", "table_scores_2_ahead.png"))

@@

@@

>>>>

```{r}
dailytruth_data %>%
  dplyr::filter(target_end_date >= min(forecast_dates$full_hub_period), 
                target_end_date <= max(forecast_dates$full_hub_period)) %>%
  ggplot(aes(y = true_value, x = target_end_date)) + 
  geom_point(size = 0.3) + 
  geom_line(size = 0.5) + 
  facet_grid(target_type ~ location_name, scales = "free") + 
  labs(y = "Observed values", x = "Date") + 
  theme_light()

@@

## Visualisation for 1, 2, 3, 4 week ahead forecasts and scores

### Plots that visualise truth vs. predictions

```{r truth-vs-forecasts, fig.height=10, fig.width=15}
# create helper function for the plot ------------------------------------------
truth_and_pred_plot <- function(df) {
  p <- df %>%
  ggplot(aes(y = true_value, x = target_end_date)) + 
  geom_linerange(aes(ymin = `0.25`, ymax = `0.75`, color = model), 
                 size = 1.5,
                 alpha = 1,
                 position = position_dodge(width = 6)) + 
  labs(y = "value", x = "date") + 
  geom_point(aes(shape = classification)) + 
  geom_line() + 
  theme_light() + 
  facet_wrap(~ location_target, scales = "free", ncol = 2) + 
  theme(legend.position = "bottom") + 
  theme(panel.background = element_rect(fill = "#e1f0fe")) + 
  scale_x_date(limits = c(min(forecast_dates$full_hub_period) + 14, 
                          max(forecast_dates$full_hub_period)))
  return(p)
}

# prepare data for plotting ----------------------------------------------------
# - load ufiltered data and pivot wider
# - add trend classification by target_end_date 
#   (and remove classification by forecast_date that is already present)

df <- unfiltered_data[quantile %in% c(0.025, 0.25, 0.75, 0.975)] %>%
  dcast(... ~ quantile, value.var = "prediction")
df <- merge(
  df[, c("classification", "speed") := NULL], 
  copy(epitrend)[, forecast_date := NULL], 
  by = c("target_end_date", "location_name", "target_type")
)

# create plots for deaths and cases --------------------------------------------

for (horizon in 1:4) {
  # plot for cases
  truth_vs_forecast_cases <- df[grepl(horizon, target)] %>%
  dplyr::filter(target_type == "case") %>%
  truth_and_pred_plot() + 
  scale_color_manual(values = brewer.pal(name = "Accent", n = 5)[2:5])

  filename <- here("analysis", "plots", 
                   paste0("truth-and-", horizon, "-wk-ahead_cases"))
  ggsave(paste0(filename, ".png"), plot = truth_vs_forecast_cases)

  saveRDS(truth_vs_forecast_cases, 
          paste0(filename, ".RDS"))

  # plot for deaths
  truth_vs_forecast_deaths <- df[grepl(horizon, target)] %>%
  dplyr::filter(target_type == "death") %>%
  truth_and_pred_plot()  + 
  scale_color_manual(values = brewer.pal(name = "Accent", n = 5)[1:5]) + 
  annotate('rect', xmin = min(forecast_dates$full_hub_period) + 14,
                xmax = max(forecast_dates$death_not_scored) + 14,
                ymin = 0,
                ymax = Inf, fill = "grey10", alpha = 0.1)

  filename <- here("analysis", "plots", 
                   paste0("truth-and-", horizon, "-wk-ahead_deaths"))
  ggsave(paste0(filename, ".png"), plot = truth_vs_forecast_deaths)

  saveRDS(truth_vs_forecast_deaths, 
          paste0(filename, ".RDS"))
}
```

### Plots that visualise the achieved scores over time

```{r scores-over-time, fig.height=10, fig.width=15}
# score forecasts --------------------------------------------------------------
# obtain scores and then do pairwise comparisons between models
summarise_by <- c("model", "horizon", "location_name", "target_type", 
                   "forecast_date", "target_end_date", "location_target", 
                   "classification")
scores_target_forecastdate <- eval_forecasts(
  data = unfiltered_data,
  summarise_by = summarise_by
)

pairwise <- pairwise_comparison(scores_target_forecastdate, 
                                summarise_by = summarise_by, 
                                baseline = "Baseline")
pairwise <- pairwise[compare_against == "Baseline"]

# create helper function to plot forecasts -------------------------------------
scores_over_time_plot <- function(scores, y = "mean_scores_ratio") {
  p <- scores %>%
  ggplot(aes(x = as.Date(target_end_date), y = get(y))) + 
  geom_line(aes(colour = model)) + 
  geom_point(aes(shape = classification, colour = model)) + 
  theme_light() + 
  facet_wrap( ~ location_target, ncol = 2, scales = "free_y") +
  theme(legend.position = "bottom") +
  labs(y = "WIS relative to baseline", x = "Date") + 
  # geom_rect(aes(xmin = min(forecast_dates$christmas) + 14,
  #               xmax = max(forecast_dates$christmas) + 14,
  #               ymin = 0,
  #               ymax = Inf), fill = "lightgrey", alpha = 0.002) +
  theme(panel.background = element_rect(fill = "#e1f0fe")) +
  scale_y_log10() + 
  scale_x_date(limits = c(min(forecast_dates$full_hub_period) + 14, 
                          max(forecast_dates$full_hub_period)))
  return(p)
}

# create plots for all horizons ------------------------------------------------
for (i in 1:4) {
  scores_over_time_case <- pairwise[horizon == i & target_type == "case"] %>%
    scores_over_time_plot() + 
    scale_color_manual(values = brewer.pal(name = "Accent", n = 5)[2:5])

  ggsave(here("analysis", "plots", paste0("score-over-time-cases-", i, ".png")), 
         plot = scores_over_time_case)

  saveRDS(scores_over_time_case, 
          here("analysis", "plots", paste0("score-over-time-cases-", i, ".RDS")))

  scores_over_time_deaths <- pairwise[horizon == i & target_type == "death"] %>%
    scores_over_time_plot() + 
    scale_color_brewer(palette = "Accent") + 
    annotate('rect', xmin = min(forecast_dates$full_hub_period) + 14,
             xmax = max(forecast_dates$death_not_scored) + 14,
             ymin = 0,
             ymax = Inf, fill = "grey10", alpha = 0.1)

  ggsave(here("analysis", "plots", paste0("score-over-time-deaths-", i, ".png")), 
         plot = scores_over_time_deaths)

  saveRDS(scores_over_time_deaths, 
          here("analysis", "plots", paste0("score-over-time-deaths-", i, ".RDS")))

}
```

### Combine the two plots

horizons <- 1:4

for (horizon in horizons) {
  truth_vs_pred_cases <- readRDS(file = here("analysis", "plots", 
                                             paste0("truth-and-", horizon, 
                                                    "-wk-ahead_cases.RDS")))

  truth_vs_pred_deaths <- readRDS(file = here("analysis", "plots", 
                                              paste0("truth-and-", horizon, 
                                                     "-wk-ahead_deaths.RDS")))

  scores_over_time_cases <- readRDS(file = here("analysis", "plots", 
                                                paste0("score-over-time-cases-", 
                                                       horizon,".RDS")))

  scores_over_time_deaths <- readRDS(file = here("analysis", "plots", 
                                                 paste0("score-over-time-deaths-", 
                                                        horizon,".RDS")))

  truth_vs_pred_cases / scores_over_time_cases / truth_vs_pred_deaths / scores_over_time_deaths + 
    plot_layout(guides = "collect") & 
    plot_annotation(tag_levels = 'A') &
    theme(legend.position = 'bottom')

  ggsave(here("analysis", "plots", paste0("figure-forecasts-", horizon, ".png")))
}

@@

@@

```{r}
# create helper function to make table
make_score_table <- function(scores, 
                             group = TRUE) {
  # copy data, change names, round column values and reduce to sign. digits
  scores <- as.data.table(scores)
  setnames(scores, 
           old = c("horizon", "model", "target_type", "interval_score", "aem", 
                   "relative_skill", "scaled_rel_skill", "sharpness", 
                   "underprediction", "overprediction", "bias", 
                   "50", "90", "location_target", "target_phase"), 
           new = c("Horizon", "Model", "Target", "WIS", "Abs. error", 
                   "rel. skill", "Skill", "Sharpness", 
                   "Underpred.", "Overpred.", "Bias", 
                   "50%-Cov.", "90%-Cov.", "Target", "Target"), 
           skip_absent = TRUE)

@@

  # remove some cols, change column order, sort table and rename horizon
  scores <- scores[, .SD,  .SDcols = !c("coverage_deviation", "rel. skill")]

  setcolorder(scores, 
              neworder = c("Target", "Horizon", "Model", "WIS"))

  scores <- scores[order(Target, Horizon)]
  scores[, Horizon := paste(Horizon, "wk ahead")]

  # calculate where to group rows
  case_rows <- nrow(scores[Target == "case"])

  table <- scores[,!c("Target")] %>%
    kable(format = "html", col.names = c(" ", names(scores)[-(1:2)])) %>% #format = "markdown"
    collapse_rows(columns = 1) %>%
    kable_styling(bootstrap_options = "striped") %>%
    column_spec(1, background = "white")

  if (group) {
    table <- table %>%
      pack_rows("Cases", 1, case_rows, indent = FALSE) %>% 
      pack_rows("Deaths", case_rows + 1, nrow(scores), indent = FALSE)
  }

  return(table)
}

### Create tables

```{r }
# create scores ----------------------------------------------------------------
# calculate coverage independently and merge with other scores
coverage <- eval_forecasts(
  filtered_data, 
  summarise_by = c("horizon", "model", "target_type", "range"), 
  compute_relative_skill = TRUE, 
  baseline = "Baseline"
)[range %in% c(50, 90), 
                   .(model, target_type, coverage, range, horizon)] %>%
  dcast(formula = ... ~ range, value.var = "coverage")

df <- eval_forecasts(
  filtered_data, 
  summarise_by = c("model", "horizon", "target_type"), 
  compute_relative_skill = TRUE, 
  baseline = "Baseline"
) %>%
  merge(coverage, all.x = TRUE, by = c("model", "horizon", "target_type"))

# tables for 1-2 weeks ahead and 3-4 weeks ahdead
table <- make_score_table(df[horizon <= 2]) 
save_kable(table, file = here("analysis", "plots", "table_scores_2_ahead.png"))

table <- make_score_table(df[horizon > 2]) 
save_kable(table, file = here("analysis", "plots", "table_scores_4_ahead.png"))

@@

### Tables for different phases of the epidemic

coverage <- eval_forecasts(
  filtered_data, 
  summarise_by = c("horizon", "model", "target_type", "target_phase", "range"), 
  compute_relative_skill = TRUE, 
  baseline = "Baseline"
)[range %in% c(50, 90), 
                   .(model, target_type, coverage, range, target_phase, horizon)] %>%
  dcast(formula = ... ~ range, value.var = "coverage")

df <- eval_forecasts(
  filtered_data, 
  summarise_by = c("model", "horizon", "target_type", "target_phase"), 
  compute_relative_skill = TRUE, 
  baseline = "Baseline"
) %>%
  merge(coverage, all.x = TRUE, by = c("model", "horizon", "target_phase", "target_type"))

tables <- list()
table_dfs <- list()
targets <- unique(df$target_phase)

horizons = 2
for (tar in targets) {
  for (i in horizons) {
    name <- paste0(tar, "-", i)

    tmp_df <- df[target_phase == tar & horizon == i, 
                 .(model, interval_score, sharpness,
                   target_phase, horizon,
                   coverage_deviation, relative_skill,  
                   underprediction, overprediction)]

    tables[[name]] <- make_score_table(tmp_df, 
                                       group = FALSE) %>%
      pack_rows(tar, 1, nrow(tmp_df), indent = FALSE)

    save_kable(tables[[name]],
               file = here("analysis", "plots",
                           paste0("table-", gsub(" ", "", name), ".png")))
  }
}

@@

library(gt)
